---
title: "Supplement: Spawner-recruit model diagnostics"
author: ""
output: 
  html_document:
    collapsed: no
    fig_caption: no
    number_sections: no
    smooth_scroll: yes
    theme: cerulean
    toc: yes
    toc_float: yes
include-in-header: |
          <style>
            h1 { font-size: 18px; }
            h2 { font-size: 12px; }
          </style>
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(ggplotify) #for as.ggplot() to help mcmc_combo() plotting
library(rstan)
library(bayesplot)
library(kableExtra)
source(here("analysis/R/data_functions.R"))

# read in model fits ---------------------------------------------------------------------
AR1.fits <- lapply(list.files(here("analysis/data/generated/model_fits/AR1"),
                              full.names = T), 
                   readRDS)
names(AR1.fits) <- unique(sp_har$CU)[order(unique(sp_har$CU))]

TVA.fits <- lapply(list.files(here("analysis/data/generated/model_fits/TVA"), 
                              full.names = T), 
                   readRDS)
names(TVA.fits) <- unique(sp_har$CU)[order(unique(sp_har$CU))]

AR1.eggs.fits <- lapply(list.files(here("analysis/data/generated/model_fits/AR1_egg_mass"),
                              full.names = T), 
                   readRDS)
names(AR1.eggs.fits) <- unique(sp_har$CU)[order(unique(sp_har$CU))]

TVA.eggs.fits <- lapply(list.files(here("analysis/data/generated/model_fits/TVA_egg_mass"),
                              full.names = T), 
                   readRDS)
names(TVA.eggs.fits) <- unique(sp_har$CU)[order(unique(sp_har$CU))]
```

This document provides diagnostic plots for several spawner-recruit models that were used to characterize Canadian-origin Yukon Chinook salmon population dynamics at the Conservation Unit scale as described in:

> Connors, B.M., O’Dell, A., Hunter, H., Glaser, D., Gill, J., Rossi, S., and Churchland, C. In review. Stock status and biological and fishery consequences of alternative harvest and rebuilding actions for Yukon River Chinook salmon (*Oncorhynchus tshawytscha*). DFO Can. Sci. Advis. Sec. Res. Doc. 2025/nnn. iv + 92 p.

Full details are provided in the document above but briefly, four state-space spawner-recruit models were fit: spawer-recruitment models with autoregressive recruitment residuals (labelled "AR1"), (2) spawner-recruitment models with time varying intrinsic productivity (labelled "TV"), (3) egg mass-recruitment models with AR1 recruitment residuals (labelled "AR1 egg mass"), and (4) egg mass-recruitment models with time varying intrinsic productivity (labelled "TV egg mass"). These models were fit to each of the nine Conservation Units for which we had data.

# Diagnostics

We fit the spawner-recruitment model in a Bayesian estimation framework with Stan (Carpenter et al. 2017; Stan Development Team 2023), which implements the No-U-Turn Hamiltonian Markov chain Monte Carlo algorithm (Hoffman and Gelman 2014)) for Bayesian statistical inference to generate a joint posterior probability distribution of all unknowns in the model. The models can be found [here](https://github.com/Pacific-salmon-assess/yukon-CK-ResDoc/tree/main/analysis/Stan).We sampled from 4 chains with 4,000 iterations each and discarded the first half as warm-up. We assessed chain convergence visually via trace plots and by ensuring that $\hat{R}$ (potential scale reduction factor; Vehtari et al. 2021) was less than 1.01 and that the effective sample size was greater than 400. Posterior predictive checks were used to make sure the model returned known values, by simulating new datasets and checking how similar they were to our observed data.

## Trace plots

These should be clearly mixed, with no single distribution deviating substantially from others (left column), and no chains exploring a strange space for a few iterations (right column).

```{r diagnostics, results = 'asis'}
# describe diagnostics in loop -----------------------------------------------------------
min_ESS_AR1 <- NULL
max_Rhat_AR1 <- NULL
min_ESS_TVA <- NULL
max_Rhat_TVA <- NULL
min_ESS_AR1_eggs <- NULL
max_Rhat_AR1_eggs <- NULL
min_ESS_TVA_eggs <- NULL
max_Rhat_TVA_eggs <- NULL

for(i in unique(sp_har$CU)){
  sub_dat <- filter(sp_har, CU==i)
  cat('\n')
  cat("### ", i, "\n") #escape loop and print CU name
  #diagnostics associated w/ AR1 model---
  sub_AR1_summary <- as.data.frame(rstan::summary(AR1.fits[[i]])$summary)
  
  sub_AR1_pars <- rstan::extract(AR1.fits[[i]])
  
  min_ESS_AR1 <- rbind(min_ESS_AR1, data.frame(CU=i, 
                                       ESS=round(min(sub_AR1_summary$n_eff, na.rm=T))))
  
  max_Rhat_AR1 <- rbind(max_Rhat_AR1, data.frame(CU=i, 
                                         Rhat=round(max(sub_AR1_summary$Rhat, na.rm=T), 3)))
  
  R <- (sub_dat$harv+sub_dat$spwn)
  R_rep <- sub_AR1_pars$H_rep[1:500,] + sub_AR1_pars$S_rep[1:500,]
  
  p <- ppc_dens_overlay(R, R_rep) +
    xlim(NA, quantile(R_rep, 0.99)) +
    theme(legend.position = "none") +
    labs(y = "density", x = "y_est", title = "AR1 posterior predictive check")
  print(p)
  
  p <- mcmc_combo(AR1.fits[[i]], pars = c("beta", "lnalpha", "sigma_R", "phi"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none()) |>
    as.ggplot() + 
    labs(title = "AR1 leading parameters")
  print(p)

  p <- mcmc_combo(AR1.fits[[i]], pars = c("D_scale", "D_sum"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "AR1 age pars")
    print(p)

  p <- mcmc_combo(AR1.fits[[i]], pars = c("Dir_alpha[1]", "Dir_alpha[2]", 
                               "Dir_alpha[3]", "Dir_alpha[4]"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "AR1 age probs")
    print(p)

  #diagnostics associated w/ TVA model---
  sub_TVA_summary <- as.data.frame(rstan::summary(TVA.fits[[i]])$summary)
  
  sub_TVA_pars <- rstan::extract(TVA.fits[[i]])
  
  min_ESS_TVA <- rbind(min_ESS_TVA, data.frame(CU=i, 
                                               ESS=round(min(sub_TVA_summary$n_eff, na.rm=T))))
  
  max_Rhat_TVA <- rbind(max_Rhat_TVA, data.frame(CU=i, 
                                                 Rhat=round(max(sub_TVA_summary$Rhat, na.rm=T), 3)))
  
  R <- (sub_dat$harv+sub_dat$spwn)
  R_rep <- sub_TVA_pars$H_rep[1:500,] + sub_TVA_pars$S_rep[1:500,]
  
  p <- ppc_dens_overlay(R, R_rep) +
    xlim(NA, quantile(R_rep, 0.99)) +
    theme(legend.position = "none") +
    labs(y = "density", x = "y_est", title = "TV posterior predictive check")
  print(p)
  
  p <- mcmc_combo(TVA.fits[[i]], pars = c("ln_alpha0", "beta", "F_rw", "mean_ln_R0"), 
                  combo = c("dens_overlay", "trace"),
                  gg_theme = legend_none()) |> 
    as.ggplot() +
    labs(title =  "TV leading parameters")
    print(p)
    
  p <- mcmc_combo(TVA.fits[[i]], pars = c("sigma_alpha", "sigma_R", "sigma_tot"), 
                  combo = c("dens_overlay", "trace"),
                  gg_theme = legend_none()) |> 
    as.ggplot() +
    labs(title =  "TV variance components")
    print(p)

  p <- mcmc_combo(TVA.fits[[i]], pars = c("D_scale", "D_sum"),
                  combo = c("dens_overlay", "trace"),
                  gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "TV age pars")
    print(p)

  p <- mcmc_combo(TVA.fits[[i]], pars = c("Dir_alpha[1]", "Dir_alpha[2]", 
                                          "Dir_alpha[3]", "Dir_alpha[4]"),
                  combo = c("dens_overlay", "trace"),
                  gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "TV age probs")
    print(p)
    
  #diagnostics associated w/ AR1 egg mass model---
  sub_AR1_eggs_summary <- as.data.frame(rstan::summary(AR1.eggs.fits[[i]])$summary)
  
  sub_AR1_eggs_pars <- rstan::extract(AR1.eggs.fits[[i]])
  
  min_ESS_AR1_eggs <- rbind(min_ESS_AR1_eggs, data.frame(CU=i, 
                                               ESS=round(min(sub_AR1_eggs_summary$n_eff, na.rm=T))))
  
  max_Rhat_AR1_eggs <- rbind(max_Rhat_AR1_eggs, data.frame(CU=i, 
                                                 Rhat=round(max(sub_AR1_eggs_summary$Rhat, na.rm=T), 3)))
  
  R <- (sub_dat$harv+sub_dat$spwn)
  R_rep <- sub_AR1_eggs_pars$H_rep[1:500,] + sub_AR1_eggs_pars$S_rep[1:500,]
  
  p <- ppc_dens_overlay(R, R_rep) +
    xlim(NA, quantile(R_rep, 0.99)) +
    theme(legend.position = "none") +
    labs(y = "density", x = "y_est", title = "AR1 egg mass posterior predictive check")
  print(p)
  
  p <- mcmc_combo(AR1.eggs.fits[[i]], pars = c("beta", "lnalpha", "sigma_R", "phi"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none()) |>
    as.ggplot() + 
    labs(title = "AR1 egg mass leading parameters")
  print(p)

  p <- mcmc_combo(AR1.eggs.fits[[i]], pars = c("D_scale", "D_sum"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "AR1 egg mass age pars")
    print(p)

  p <- mcmc_combo(AR1.eggs.fits[[i]], pars = c("Dir_alpha[1]", "Dir_alpha[2]", 
                               "Dir_alpha[3]", "Dir_alpha[4]"),
             combo = c("dens_overlay", "trace"),
             gg_theme = legend_none())|> 
    as.ggplot() +
    labs(title = "AR1 egg mass age probs")
    print(p)    
  
#diagnostics associated w/ TVA egg mass model---
sub_TVA_eggs_summary <- as.data.frame(rstan::summary(TVA.eggs.fits[[i]])$summary)

sub_TVA_eggs_pars <- rstan::extract(TVA.eggs.fits[[i]])

min_ESS_TVA_eggs <- rbind(min_ESS_TVA_eggs, data.frame(CU=i, 
                                                       ESS=round(min(sub_TVA_eggs_summary$n_eff, na.rm=T))))

max_Rhat_TVA_eggs <- rbind(max_Rhat_TVA_eggs, data.frame(CU=i, 
                                                         Rhat=round(max(sub_TVA_eggs_summary$Rhat, na.rm=T), 3)))

R <- (sub_dat$harv+sub_dat$spwn)
R_rep <- sub_TVA_eggs_pars$H_rep[1:500,] + sub_TVA_eggs_pars$S_rep[1:500,]

p <- ppc_dens_overlay(R, R_rep) +
  xlim(NA, quantile(R_rep, 0.99)) +
  theme(legend.position = "none") +
  labs(y = "density", x = "y_est", title = "TVA egg mass posterior predictive check")
print(p)

p <- mcmc_combo(TVA.eggs.fits[[i]], pars = c("ln_alpha0", "beta", "F_rw", "mean_ln_R0"), 
                combo = c("dens_overlay", "trace"),
                gg_theme = legend_none()) |> 
  as.ggplot() +
  labs(title =  "TV egg mass leading parameters")
print(p)

p <- mcmc_combo(TVA.eggs.fits[[i]], pars = c("sigma_alpha", "sigma_R", "sigma_tot"), 
                combo = c("dens_overlay", "trace"),
                gg_theme = legend_none()) |> 
  as.ggplot() +
  labs(title =  "TV egg mass variance components")
print(p)

p <- mcmc_combo(TVA.eggs.fits[[i]], pars = c("D_scale", "D_sum"),
                combo = c("dens_overlay", "trace"),
                gg_theme = legend_none())|> 
  as.ggplot() +
  labs(title = "TV egg mass age pars")
print(p)

p <- mcmc_combo(TVA.eggs.fits[[i]], pars = c("Dir_alpha[1]", "Dir_alpha[2]", 
                                        "Dir_alpha[3]", "Dir_alpha[4]"),
                combo = c("dens_overlay", "trace"),
                gg_theme = legend_none())|> 
  as.ggplot() +
  labs(title = "TV egg mass age probs")
print(p)
      cat('\n')
}
```

## ESS and $\hat{R}$

We aimed for minimum effective sample sizes that are greater than ***2000*** and $\hat{R}$ values less than 1.05.

### AR1 model:

```{r AR1-diag}
AR1.diag <- left_join(min_ESS_AR1, max_Rhat_AR1, by = "CU")
kable(AR1.diag)
```

### Time varying (TV) model:

```{r TV-diag}
TVA.diag <- left_join(min_ESS_TVA, max_Rhat_TVA, by = "CU")
kable(TVA.diag)
```

### Egg mass AR1 model:

```{r AR1egg-diag}
AR1egg.diag <- left_join(min_ESS_AR1_eggs, max_Rhat_AR1_eggs, by = "CU")
kable(AR1egg.diag)
```

### Time varying (TV) egg mass model:

```{r TVAegg-diag}
AR1egg.diag <- left_join(min_ESS_TVA_eggs, max_Rhat_TVA_eggs, by = "CU")
kable(AR1egg.diag)
```

